---
title: 'BioData Club/Tidy Tuesday: African American Accomplishments'
author: "Pierrette Lo"
date: "10/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggiraph)
library(ggbump)
library(PNWColors)
library(tidytext)
library(igraph)
library(networkD3)
```

Read in data

```{r}
firsts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-06-09/firsts.csv')

science <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-06-09/science.csv')
```

```{r}
skimr::skim(firsts)
```


## Firsts since 2000 

(kind of cheating a little - chose a small enough dataset that I could visually inspect for problems/patterns)

```{r}
firsts_2000 <- firsts %>%
  filter(year >= 2000) %>% 
  mutate(category = as.factor(category)) %>%  
  #complete(year, category) %>% 
  group_by(year, category) %>% 
  mutate(id = row_number(person)) %>% 
  select(year, category, id, everything()) %>% 
  ungroup()

# category has 8 levels
levels(firsts_2000$category)

```


Data cleanup:
- remove "(See also" and "[*]" in `person`
- replace "First African-American" in `accomplishment` to "..."
- add tooltip column for ggiraph (use `htmlEscape` to sanitize text for use as attribute, see ref: https://github.com/davidgohel/ggiraph/issues/5)

The `gender` column is ambiguous - there are many obviously female `person`s who don't have "female" in `gender`. The data dictionary also isn't clear - it says "Gender - indicates either female AND African-American, or a more general African-American first". For this exercise, I'm going to assume that this means "first African-American AND first woman of any race".

```{r}
firsts_2000_clean <- firsts_2000 %>% 
  mutate(person = str_replace(person, "\\(See also", "")) %>% 
  mutate(person = str_replace_all(person, "\\[.*\\]", ""),
         accomplishment = str_replace(accomplishment, "First African-American", "...")) %>% 
  mutate(tooltip = htmltools::htmlEscape(paste0(accomplishment, ":\n", person), TRUE))
```

Static Plot

```{r, fig.height=8, fig.width=8}
firsts_2000_clean %>% 
  ggplot(aes(x = year, y = id, fill = gender)) +
  geom_point(size = 3,
             shape = 21,
             color = "grey15") +
  facet_wrap(~category, nrow = 8) +
  xlab(NULL) +
  ylab(NULL) +
  scale_fill_manual(values = c("#1380A1", "#FAAB18"),
                     name = NULL,
                       labels = c("First AA", "First AA & first woman of any race")) +
  xlim(1999, 2021) +
  ylim(0, 5.5) +
  theme_dark() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_line(color = "grey10"),
        axis.text.x = element_text(color = "grey10"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_line(color = "#cbcbcb",
                                          linetype = "dotted"),
        panel.grid.minor.x = element_line(color = "#cbcbcb",
                                          linetype = "dotted"),
        legend.position = "top",
        legend.justification = "left",
        legend.background = element_rect(fill = "grey50"),
        plot.background = element_rect(fill = "grey50")) +
  ggtitle("2000-2019: The First African-American...")
```

Interactive plot:

```{r}
p <- firsts_2000_clean %>% 
  ggplot(aes(x = year, y = id, fill = gender)) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = tooltip),
                         size = 3, shape = 21, color = "grey15") +
  facet_wrap(~category, nrow = 8) +
  xlab(NULL) +
  ylab(NULL) +
  scale_fill_manual(values = c("#1380A1", "#FAAB18"),
                     name = NULL,
                       labels = c("First AA", "First AA & first woman of any race")) +
  xlim(1999, 2021) +
  ylim(0, 5.5) +
  theme_dark() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.x = element_line(color = "grey10"),
        axis.text.x = element_text(color = "grey10"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_line(color = "#cbcbcb",
                                          linetype = "dotted"),
        panel.grid.minor.x = element_line(color = "#cbcbcb",
                                          linetype = "dotted"),
        legend.position = "top",
        legend.justification = "left",
        legend.background = element_rect(fill = "grey50"),
        plot.background = element_rect(fill = "grey50")) +
  ggtitle("2000-2019: The First African-American...")

girafe(code = print(p),
       options = list(opts_hover(css = "stroke:darkorange;"),
                      opts_hover_inv(css = "opacity:0.5;"),
                      opts_tooltip(opacity = 0.5),
                      opts_sizing(rescale = TRUE, width = 0.7))
)
```

## Bump plot - number of firsts per category over the years


```{r}
bump_plot_data <- firsts %>% 
  mutate(century = year - year %% 100) %>% 
  group_by(century, category) %>% 
  summarize(n_firsts = n()) %>% 
  group_by(century) %>% 
  mutate(rank = rank(-n_firsts, ties.method = "random")) %>% 
  ungroup()
```


```{r}
bump_plot_data %>% 
  ggplot(aes(x = century, y = rank, color = category)) +
  geom_bump(size = 1) +
  geom_point(size = 4) +
  geom_text(data = bump_plot_data %>% filter(century == min(century)),
            aes(x = century - 10, label = category),
            hjust = 1) +
  geom_text(data = bump_plot_data %>% filter(century == max(century)),
            aes(x = century + 10, label = category),
            hjust = 0) +
  scale_y_reverse() +
  scale_x_continuous(limits = c(1600, 2100),
                     breaks = c(1700, 1800, 1900, 2000)) +
  scale_color_manual(values = pnw_palette("Bay", 8)) +
    labs(title = "African-American accomplishments, ranked by number of firsts per category",
       x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.background = element_rect(fill = "black"),
        panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(color = "white"),
        plot.title = element_text(color = "white", hjust = 0.5))

ggsave("aaa_bump_plot.png", width = 7, height = 5)
```

## Network of occupations

```{r}
sci_tokens <- science %>%
  select(name, occupation_s) %>%
  # replace "and" with ";" + delete asides + replace weird entries
  mutate(occupation_s = str_replace(occupation_s, "and", ";"),
         occupation_s = str_replace(occupation_s, "\\[.*\\]", ""),
         occupation_s = str_replace(occupation_s, "\\(.*\\)", ""),
         occupation_s = str_replace(occupation_s, "Woods Hole Marine Biology Institute biologist", "biologist"),
         occupation_s = str_replace(occupation_s, "ZoologistexplorerAnthropologist", "zoologist; explorer; anthropologist")) %>% 
  unnest_tokens(job, occupation_s, token = "regex", pattern = ";", drop = FALSE) %>% 
  filter(job != " ") %>% 
  arrange(job)
```

How many unique occupations?
```{r}
n_distinct(sci_tokens$job)
```


Next: make igraph and format for networkD3 use

```{r}
sci_simple <- sci_tokens %>%
  left_join(sci_tokens, by = c("name", "occupation_s")) %>% 
  filter(job.x < job.y) %>% 
  group_by(job.x, job.y) %>% 
  summarize(weight = n()) %>% 
  ungroup() 
  
simpleNetwork(sci_simple)
```


```{r}

sci_igraph <- sci_simple %>%
  graph_from_data_frame(directed = FALSE)

# find community structure using short random walks
wc <- cluster_walktrap(sci_igraph)
members <- membership(wc)

sci_d3 <- igraph_to_networkD3(sci_igraph, group = members)

forceNetwork(Links = sci_d3$links,
             Nodes = sci_d3$nodes,
             Source = "source",
             Target = "target",
             NodeID = "name",
             Group = "group",
             opacity = 0.8,
             opacityNoHover = 0.8,
             bounded = TRUE,
             fontSize = 12,
             charge = -40)
```

Problem - multiple clusters involving "inventor" because it can be in job.x or job.y -> need to figure out how to make order not matter

Try next:
- flipbook!
- animation - number of accomplishments by category, per year - racing bar chart
- wordcloud
- annotate timeline with historic milestones/bands of time for wars, etc.
- use {forcats} to collapse categories e.g. for occupations
- pull pictures from wikipedia
- ggbump top 5 occupations/categories
- network graph of occupations that are related

